# RNA-Seq (an example)

## Introduction


This workflow below is based on @Pipperrana. For analysis, we used the **DESeq2** package [@Love2014] and also considered @anrlove2020 [description](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html).

The research questions and databases used in the workflow are based on the publication of @Gerarduzzi2017, RNA sequencing data can be downloaded from the Gene Expression Omnibus (GEO) database (GEO accession: [GSE85209] (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE85209)).

## Loading packages

```{r}
# Load library for DESeq2
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# .libPaths()
# BiocManager::install("DESeq2")
# devtools::install_github("stephenturner/annotables")

library(DESeq2)

# Load library for RColorBrewer
library(RColorBrewer)

# Load library for pheatmap
# install.packages("pheatmap")
library(pheatmap)

# Load library for tidyverse
library(tidyverse)
```


## Read counts

```{r}
smoc2_rawcounts <- read.csv("data/fibrosis_smoc2_rawcounts_unordered.csv") 
rownames(smoc2_rawcounts) <- smoc2_rawcounts$X
smoc2_rawcounts$X <- NULL

# Explore the first six observations of smoc2_rawcounts
head(smoc2_rawcounts)

# Explore the structure of smoc2_rawcounts
str(smoc2_rawcounts)
```

## Distribution of counts

```{r}
library(ggplot2)
ggplot(smoc2_rawcounts) + 
  geom_histogram(aes(x=smoc2_fibrosis1), bins = 300)
summary(smoc2_rawcounts$smoc2_fibrosis1)
```

## Setting of metadata

```{r}
# Create genotype vector
genotype <- c("smoc2_oe", "smoc2_oe", "smoc2_oe", "smoc2_oe", "smoc2_oe", "smoc2_oe", "smoc2_oe")

# Create condition vector
condition <- c("fibrosis", "fibrosis", "fibrosis", "fibrosis", "normal", "normal", "normal")

# Create data frame
smoc2_metadata <- data.frame(genotype, condition)

# Assign the row names of the data frame
rownames(smoc2_metadata) <- c("smoc2_fibrosis1", "smoc2_fibrosis2", "smoc2_fibrosis3", "smoc2_fibrosis4", "smoc2_normal1", "smoc2_normal3", "smoc2_normal4")

smoc2_metadata
```

## Matching metadata and counts data

```{r}
# Use the match() function to reorder the columns of the raw counts
reorder_idx <- match(rownames(smoc2_metadata), colnames(smoc2_rawcounts))

# Reorder the columns of the count data
reordered_smoc2_rawcounts <- smoc2_rawcounts[ , reorder_idx]

all(rownames(smoc2_metadata) == colnames(reordered_smoc2_rawcounts))

# Create a DESeq2 object
dds_smoc2 <- DESeqDataSetFromMatrix(countData = reordered_smoc2_rawcounts,
                              colData = smoc2_metadata,
                              design = ~ condition)
```

## Normalizing counts

```{r}
# Determine the size factors to use for normalization
dds_smoc2 <- estimateSizeFactors(dds_smoc2)

sizeFactors(dds_smoc2)

# Extract the normalized counts
smoc2_normalized_counts <- counts(dds_smoc2, normalized=TRUE)
# smoc2_normalized_counts
```


## Quality control - Hierarchical heatmap

```{r}
# Transform the normalized counts 
vsd_smoc2 <- vst(dds_smoc2, blind = TRUE)

# Extract the matrix of transformed counts
vsd_mat_smoc2 <- assay(vsd_smoc2)

# Compute the correlation values between samples
vsd_cor_smoc2 <- cor(vsd_mat_smoc2) 

# Plot the heatmap
pheatmap(vsd_cor_smoc2, annotation = select(smoc2_metadata, condition))
```

## Quality control - Principal component analysis

```{r}
# Transform the normalized counts 
vsd_smoc2 <- vst(dds_smoc2, blind = TRUE)

# Plot the PCA of PC1 and PC2
plotPCA(vsd_smoc2, intgroup="condition")
```


## Creating the DeSeq2 object

```{r}
# Create DESeq2 object
dds_smoc2 <- DESeqDataSetFromMatrix(countData = reordered_smoc2_rawcounts,
                              colData = smoc2_metadata,
                              design = ~ condition)

# Run the DESeq2 analysis
dds_smoc2 <- DESeq(dds_smoc2)
```


## DESeq2 model - dispersion


```{r}
mean_counts <- apply(smoc2_rawcounts, 1, mean)
variance_counts <- apply(smoc2_rawcounts, 1, var)
df <- data.frame(mean_counts, variance_counts)
ggplot(df) + geom_point(aes(x=mean_counts, y=variance_counts)) + 
  scale_y_log10() + 
  scale_x_log10() +
  labs(x="Mean counts per gene", y="Vaiance per gene") 
```

Dispersion formula: $Var=\mu+\alpha\times\mu^2$

- $Var$: variance
- $\mu$: mean
- $\alpha$: dispersion

```{r}
# Plot dispersions
plotDispEsts(dds_smoc2)
```


## DESeq2 model - contrasts


DESeq2 Negative Binomial Model:

$$K_{ij} \sim NB(\mu_{ij},\alpha_i)$$
$$\mu_{ij}=s_{j}q_{ij}$$


$$log_2(q_{ij})=x_j\beta_j$$


- $K_{ij}$ - raw count for gene i, samle j
- $s_{ij}$ - size factor
- $g_{ij}$ - normalized count


```{r}
# Extract the results of the differential expression analysis
smoc2_res <- results(dds_smoc2, 
                contrast = c("condition", "fibrosis", "normal"), 
                alpha = 0.05)
plotMA(smoc2_res)
```



### DESeq2 rsults - LFC shrinkage

```{r}
# Shrink the log2 fold change estimates to be more accurate
smoc2_res <- lfcShrink(dds_smoc2,
                    contrast = c("condition", "fibrosis", "normal"),
                    res = smoc2_res, type = "ashr")
plotMA(smoc2_res)
```


## DESeq2 results

```{r}
mcols(smoc2_res)
head(smoc2_res)
summary(smoc2_res)
```


```{r}
# Extract results
smoc2_res <- results(dds_smoc2, 
                contrast = c("condition", "fibrosis", "normal"), 
                alpha = 0.05, 
                lfcThreshold = 0.32)

# Shrink the log2 fold changes
smoc2_res <- lfcShrink(dds_smoc2, 
                    contrast = c("condition", "fibrosis", "normal"), 
                    res = smoc2_res, type = "ashr")
```


```{r}
# Get an overview of the results    
summary(smoc2_res)
```


## DESeq2 significant results

```{r}
# Save results as a data frame
smoc2_res_all <- data.frame(smoc2_res)

# Subset the results to only return the significant genes with p-adjusted values less than 0.05
smoc2_res_sig <- subset(smoc2_res_all, padj < 0.05)
head(smoc2_res_sig)
```


## Visualization of results - MA és Vulcano plots

```{r}
# Create MA plot
plotMA(smoc2_res)

# Generate logical column 
smoc2_res_all <- data.frame(smoc2_res) %>% mutate(threshold = padj < 0.05)
              
# Create the volcano plot
ggplot(smoc2_res_all) + 
        geom_point(aes(x = log2FoldChange, y = -log10(padj), color = threshold)) + 
        xlab("log2 fold change") + 
        ylab("-log10 adjusted p-value") + 
        theme(legend.position = "none", 
              plot.title = element_text(size = rel(1.5), hjust = 0.5), 
              axis.title = element_text(size = rel(1.25)))
```


## Visualization of results -  heatmap

```{r}
# Subset normalized counts to significant genes
sig_norm_counts_smoc2 <- smoc2_normalized_counts[rownames(smoc2_res_sig), ]

# Choose heatmap color palette
heat_colors <- brewer.pal(n = 6, name = "YlOrRd")

# Plot heatmap
pheatmap(sig_norm_counts_smoc2, 
         color = heat_colors, 
         cluster_rows = TRUE, 
         show_rownames = FALSE,
         annotation = select(smoc2_metadata, condition), 
         scale = "row")
```


## Significant genes

```{r}
library(annotables)
head(grcm38)

smoc2_res_all <- data.frame(smoc2_res) %>% 
  rownames_to_column(var="ensgene") %>% 
  left_join(, y=grcm38[, c("ensgene", "symbol", "description")],
            by="ensgene")

head(smoc2_res_all)

smoc2_res_sig <- subset(smoc2_res_all, padj < 0.05)
smoc2_res_sig <- smoc2_res_sig %>% arrange(padj)
head(smoc2_res_sig)
```



##  Top 20 genes

```{r}
top_20 <- data.frame(sig_norm_counts_smoc2)[1:20, ] %>% 
  rownames_to_column(var="ensgene")

top_20 <- top_20 %>% pivot_longer(cols = 2:8, values_to = "normalized_counts", names_to="samplename")

head(top_20)

top_20 <- inner_join(top_20, 
                     rownames_to_column(smoc2_metadata, var="samplename"),
                     by="samplename")
# write.table(x = top_20, file = "tmp.csv", sep=";", dec=",", row.names = F, quote = T)
ggplot(top_20) + 
  geom_point(aes(x=ensgene, y=normalized_counts, color=condition)) +
  scale_y_log10() +
  labs(title = "Top 20 genes", x="Genes", y="Normalized counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1))

```



