# Summary

## RMarkdown

With R markdown, it is easy to reproduce not only the analysis used, but also the entire report. The advantage of using R markdown (versus a script) is that you can combine computation with explanation. In other words, you can weave the outputs of your R code, like figures and tables, with text to create a report.

|                 | RMarkdown                            | R script          |
|-----------------|--------------------------------------|-------------------|
| File extension  | .Rmd                                 | .R                |
| File contents   | R code + Markdown text + YAML header | R code            |
| Reproducibility | analysis + entire report             | only the analysis |
| Output format   | PDF, HTML, Word DOCX                 | \-                |

## Advanced data manipulation

+---------------------------+--------------------------------------------------------+------------------------------------------------+
|                           | Base R                                                 | Tidyverse R                                    |
+===========================+========================================================+================================================+
| Most used function        | `[]`                                                   | `%>%`                                          |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Import                    | `read.table()` `rio::import()`                         | `readr::read_delim()`                          |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-import)      | [Link](tidyverse-r.html#tr-import)             |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Export                    | `write.table()` `rio::export()`                        | `readr::write_delim()`                         |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-export)      | [Link](tidyverse-r.html#tr-export)             |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Inspecting dataset        | `str()`                                                | `dplyr::glimpse()`                             |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-inspect)     | [Link](tidyverse-r.html#tr-inspect)            |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Working with factors      | `factor()`                                             | `forcats::fct_infreq()`                        |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-factor)      | [Link](tidyverse-r.html#tr-factor)             |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Working with strings      | `paste()`                                              | `stringr::str_c()`                             |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-string)      | [Link](tidyverse-r.html#tr-string)             |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Working with column names | `names()`                                              | `dplyr::rename()`                              |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-col-names)   | [Link](tidyverse-r.html#tr-col-names)          |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Working with row names    | `rownames()`                                           | `tibble::rowid_to_column()`                    |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-row-names)   | [Link](tidyverse-r.html#tr-row-names)          |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Filtering columns         | `[]`                                                   | `dplyr::select()`                              |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-filter-cols) | [Link](tidyverse-r.html#tr-filter-cols)        |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Filtering rows            | `[]`                                                   | `dplyr::filter()`                              |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-filter-rows) | [Link](tidyverse-r.html#tr-filter-rows)        |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Sorting rows              | `[]`                                                   | `dplyr::arrange()`                             |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-sort)        | [Link](tidyverse-r.html#tr-sort)               |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Changing your data        | `cut()`                                                | `dplyr::mutate()`                              |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-changing)    | [Link](tidyverse-r.html#tr-changing)           |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Summarising data          | `aggregate()`                                          | `dplyr::summarise()`                           |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-summary)     | [Link](tidyverse-r.html#tr-summary)            |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Combining datasets        | `merge()`                                              | `dplyr::left_join()`                           |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-joins)       | [Link](tidyverse-r.html#tr-joins)              |
+---------------------------+--------------------------------------------------------+------------------------------------------------+
| Reshaping data            | `reshape2::melt()` `reshape2::dcast()`                 | `tidyr::pivot_longer()` `tidyr::pivot_wider()` |
|                           |                                                        |                                                |
|                           | [Link](advanced-data-manipulation.html#br-reshape)     | [Link](tidyverse-r.html#tr-reshape)            |
+---------------------------+--------------------------------------------------------+------------------------------------------------+

## Modern graphics in R - ggplot2

### The grammar of graphics

The grammar of graphics lies at the heart of ggplot2 and also lies at the heart of how we define our data visualizations [@10.5555/1088896].

| Component  | Description                                 |
|------------|---------------------------------------------|
| Data       | Raw data that we'd like to visualize        |
| Geometries | Shapes that we use to visualize             |
| Aesthetics | Properties of geometries (size, color etc.) |
| Mapping    | Mapping between data and aesthetics         |

: (#tab:gog) The Grammar of Graphics

```{r}
library(tidyverse)
# a tibble for data, 3 rows, 4 columns
d.tbl <- tribble(
  ~group, ~score.1, ~score.2, ~score.3,
  "AA", 15, 42, 12, 
  "BB", 20, 28, 18,
  "CC", 35, 12, 21
)
```


```{r fig.align="center", fig.width=4, fig.height=3.3}
# Scatterplot
#   Data: d.tbl
#   Geometry: point
#   Aesthetics: x, y
#   Mapping: x=score.1, y=score.2
ggplot(data=d.tbl, mapping=aes(x=score.1, y=score.2)) + geom_point()
# Column Graph
#   Data: d.tbl
#   Geometry: column
#   Aesthetics: x, y
#   Mapping: x=score.1, y=score.2
ggplot(data=d.tbl, mapping=aes(x=score.1, y=score.2)) + geom_col()
# Line Graph
#   Data: d.tbl
#   Geometry: line
#   Aesthetics: x, y
#   Mapping: x=score.1, y=score.2
ggplot(data=d.tbl, mapping=aes(x=score.1, y=score.2)) + geom_line()
# all in one
ggplot(data=d.tbl, mapping=aes(x=score.1, y=score.2)) + 
  geom_point() + geom_col() + geom_line()
# Scatterplot
#   Data: d.tbl
#   Geometry: point
#   Aesthetics: x, y, size, color
#   Mapping: x=score.1, y=score.2, size=score.3, color=group
ggplot(data=d.tbl, 
       mapping=aes(x=score.1, y=score.2, size=score.3, color=group)) + 
  geom_point()
# Column Graph
#   Data: d.tbl
#   Geometry: column
#   Aesthetics: x, y, fill
#   Mapping: x=score.1, y=score.2, fill=score.3
ggplot(data=d.tbl, mapping=aes(x=score.1, y=score.2, fill=group)) + 
  geom_col()
```

### Geometries with required and optional aesthetics.

+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| Geometry           | Required aesthetics                              | Optional aesthetics                                                                     |
+====================+==================================================+=========================================================================================+
| `geom_abline()`    | `slope`, `intercept`                             | `alpha`, `color`, `linetype`, `size`                                                    |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_hline()`     | `yintercept`                                     | `alpha`, `color`, `linetype`, `size`                                                    |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_vline()`     | `xintercept`                                     | `alpha`, `color`, `linetype`, `size`                                                    |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_area()`      | `x`, `ymin`, `ymax`                              | `alpha`, `colour`, `fill`, `group`, `linetype`, `size`                                  |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_col()`       | `x`, `y`                                         | `alpha`, `colour`, `fill`, `group`, `linetype`, `size`                                  |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_bar()`       | `x`, `y`                                         | `alpha`, `colour`, `fill`, `group`, `linetype`, `size`                                  |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_boxplot()`   | `x`, `lower`, `middle`, `upper`, `ymax`, `ymin`) | `alpha`, `color`, `fill`, `group`, `linetype`, `shape`, `size`, `weight`                |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_density()`   | `x`, `y`                                         | `alpha`, `color`, `fill`, `group`, `linetype`, `size`, `weight`                         |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_dotplot()`   | `x`, `y`                                         | `alpha`, `color`, `fill`, `group`, `linetype`, `stroke`                                 |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_histogram()` | `x`                                              | `alpha`, `color`, `fill`, `linetype`, `size`, `weight`                                  |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_jitter()`    | `x`, `y`                                         | `alpha`, `color`, `fill`, `shape`, `size`                                               |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_line()`      | `x`, `y`                                         | `alpha`, `color`, `linetype`, `size`                                                    |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_point()`     | `x`, `y`                                         | `alpha`, `color`, `fill`, `shape`, `size`                                               |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_ribbon()`    | `x`, `ymax`, `ymin`                              | `alpha`, `color`, `fill`, `linetype`, `size`                                            |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_smooth()`    | `x`, `y`                                         | `alpha`, `color`, `fill`, `linetype`, `size`, `weight`                                  |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+
| `geom_text()`      | `label`, `x`, `y`                                | `alpha`, `angle`, `color`, `family`, `fontface`, `hjust`, `lineheight`, `size`, `vjust` |
+--------------------+--------------------------------------------------+-----------------------------------------------------------------------------------------+


### Plot types

```{r gemabraklistaja, echo=FALSE, fig.align="center", fig.width=7, fig.height=19, fig.cap="Plot types and geometries"}
library(gridExtra)
library(ggthemes)
data(survey, package = "MASS")

survey$Smoke <- factor(survey$Smoke, levels=c("Never", "Occas", "Regul", "Heavy"))
survey$Exer <- factor(survey$Exer, levels=c("None", "Some", "Freq"))

p1 <- ggplot(na.omit(survey), aes(x=Smoke, y=Pulse, fill=Smoke)) + geom_boxplot() + labs(title="Boxplot", subtitle = "geom_boxplot()") + theme_fivethirtyeight() + theme(legend.position = "none")

p2 <- ggplot(na.omit(survey), aes(x=Smoke, y=Pulse, fill=Smoke)) + geom_violin() + labs(title="Violinplot", subtitle = "geom_violin()") + theme_fivethirtyeight() + theme(legend.position = "none")

p3 <- ggplot(na.omit(survey), aes(x=Smoke, y=Pulse, colour=Smoke)) + geom_jitter() + labs(title="Stripchart", subtitle = "geom_jitter()") + theme_fivethirtyeight() + theme(legend.position = "none")

p4 <- ggplot(na.omit(survey), aes(x=Pulse, fill=Smoke)) + geom_dotplot(binwidth = 2.5) + labs(title="Dotplot", subtitle = "geom_dotplot()") + theme_fivethirtyeight() + theme(legend.position = "none")

p5 <- ggplot(na.omit(survey), aes(x=Smoke, fill=Smoke)) + geom_bar() + labs(title="Column graph", subtitle = "geom_bar() or geom_col()") + theme_fivethirtyeight() + theme(legend.position = "none")

p6 <- ggplot(na.omit(survey), aes(x=Height, y=Wr.Hnd, colour=Sex)) + geom_point() + geom_smooth(se = F, method = "lm") + labs(title="Scatterplot", subtitle = "geom_point() and geom_smooth()") + theme_fivethirtyeight() + theme(legend.position = "none")

p7 <- ggplot(data=survey, aes(x=Height)) + 
        geom_histogram( binwidth=4, colour = "blue", fill = "lightblue") + labs(title="Histogram", subtitle = "geom_histogram()") + theme_fivethirtyeight() + theme(legend.position = "none")

p8 <- ggplot(data=survey, aes(x=Height, colour=Exer, linetype=Sex)) + 
        geom_freqpoly(binwidth=4, size=1.2) + labs(title="Frequency polygon", subtitle = "geom_freqpoly()") + theme_fivethirtyeight() + theme(legend.position = "none")

p9 <- ggplot(data=survey, aes(x=Height, fill = Exer)) + geom_density(alpha=0.2, size=0.7) + labs(title="Smoothed histogram", subtitle = "geom_density()") + theme_fivethirtyeight() + theme(legend.position = "none")

p10 <- ggplot(data=survey[!is.na(survey$Exer) & !is.na(survey$Sex),], 
             aes(x=Exer, y=Pulse, fill=Sex)) + 
        stat_summary(fun=mean, geom="point", position=position_dodge(width=0.90), colour="black") + 
        stat_summary(fun.data=mean_cl_normal, geom="errorbar", 
                     position=position_dodge(width=0.90), width=0.11) +
        scale_fill_manual("Sex",values = c("Female"="#bccdff", "Male"="#81a1ff")) + 
        coord_cartesian(ylim = c(30, 100)) + labs(title="Errorbar", subtitle = "geom_error() and geom_point()") + theme_fivethirtyeight() + theme(legend.position = "none")


# Lollipop ---------------------
cty_mpg <- aggregate(mpg$cty, by=list(mpg$manufacturer), FUN=mean)  # aggregate
colnames(cty_mpg) <- c("make", "mileage")  # change column names
cty_mpg <- cty_mpg[order(cty_mpg$mileage), ]  # sort
cty_mpg$make <- factor(cty_mpg$make, levels = cty_mpg$make)  # to retain the order in plot.
p11 <- ggplot(cty_mpg, aes(x=make, y=mileage)) + 
  geom_point(size=3) + 
  geom_segment(aes(x=make, 
                   xend=make, 
                   y=0, 
                   yend=mileage))  + labs(title="Lollipop", subtitle = "geom_segment() and geom_point()") + theme_fivethirtyeight() + theme(legend.position = "none") +  theme(axis.text.x = element_text(angle=65, vjust=0.6))



library(gcookbook) # Load gcookbook for the climate data set
library(dplyr)

# Grab a subset of the climate data
climate_mod <- climate %>%
  filter(Source == "Berkeley") %>%
  select(Year, Anomaly10y, Unc10y)

# With a dotted line for upper and lower bounds
p12 <- ggplot(climate_mod, aes(x = Year, y = Anomaly10y)) +
  geom_line(aes(y = Anomaly10y - Unc10y), colour = "grey50", linetype = "dotted") +
  geom_line(aes(y = Anomaly10y + Unc10y), colour = "grey50", linetype = "dotted") +
  geom_line() + labs(title="Line graph", subtitle = "geom_line()") + theme_fivethirtyeight() + theme(legend.position = "none")


library(gcookbook)  # Load gcookbook for the uspopage data set

# Create the base plot
p13 <- ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area() +  labs(title="Area plot", subtitle = "geom_area()") + theme_fivethirtyeight() + theme(legend.position = "none")


p14 <- ggplot(data=survey, aes(sample=Height, colour=Sex)) + stat_qq() + labs(title="QQ-plot", subtitle = "stat_qq() and stat_qq_line()") + theme_fivethirtyeight() + theme(legend.position = "none") + stat_qq_line()

p15 <- ggplot(data=survey, aes(x=Height)) + stat_ecdf(geom = "step") + labs(title="Distribution", subtitle = "stat_ecdf()") + theme_fivethirtyeight() + theme(legend.position = "none")


grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, p13, p14, p15, ncol=2)
```


### Themes

```{r fig.align="center", fig.width=6, fig.height=8, echo=FALSE}
novekedes <- read.table(file = "data/growth.csv", sep=",", header = T, dec = ".", 
                        stringsAsFactors = T)
p0  <- ggplot(data=novekedes, aes(x=diet, y=gain)) + geom_boxplot()
p1  <- p0 + theme_gray() + labs(subtitle = "theme_gray()")
p2  <- p0 + theme_bw() + labs(subtitle = "theme_bw()")
p3  <- p0 + theme_classic() + labs(subtitle = "theme_classic()")
p4  <- p0 + theme_light() + labs(subtitle = "theme_light()")
p5  <- p0 + theme_minimal() + labs(subtitle = "theme_minimal()")
p6  <- p0 + theme_clean() + labs(subtitle = "theme_clean()")
p7  <- p0 + theme_economist() + labs(subtitle = "theme_economist()")
p8  <- p0 + theme_economist_white() + labs(subtitle = "theme_economist_white()")
p9  <- p0 + theme_fivethirtyeight() + labs(subtitle = "theme_fivethirtyeight()")
p10 <- p0 + theme_stata() + labs(subtitle = "theme_stata()")
p11 <- p0 + theme_tufte() + labs(subtitle = "theme_tufte()")
p12 <- p0 + theme_pander() + labs(subtitle = "theme_pander()")
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, nrow=4, ncol=3)
```

