---
title: "Warm-up exercise"
subtitle: "A solution"
author: "Kálmán Abari"
date: "`r Sys.Date()`"
output: tint::tintHtml
bibliography: skeleton.bib
link-citations: yes
nocite: '@*'
---

```{r setup, include=FALSE}
library(tint)
# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      cache.extra = packageVersion('tint'), 
                      message = F, 
                      tidy = T, 
                      collapse = T, 
                      warning = F, 
                      comment = "#>")
# example R options set globally
options(width = 60)
options(htmltools.dir.version = FALSE)
```

# Data structures

## Problems

Consider the following set of attributes about the American Film Institute's top-five movies ever from their 2007 list.

1.  What code would you use to create a vector named `Movie` with the values `Citizen Kane`, `The Godfather`, `Casablanca`, `Raging Bull`, and `Singing in the Rain`? (Hints: `object <- c()`, Working with character in R)

```{r}
# Solution ----
Movie <- c("Citizen Kane", "The Godfather", "Casablanca", "Raging Bull", "Singing in the Rain")
Movie
```

2.  What code would you use to create a vector — giving the year that the movies in Problem 1 were made — named `Year` with the values 1941, 1972, 1942, 1980, and 1952? 

```{r}
# Solution ----
Year <- c(1941, 1972, 1942, 1980, 1952)
Year
```

3.  What code would you use to create a vector — giving the run times in minutes of the movies in Problem 1 — named `RunTime` with the values 119, 177, 102, 129, and 103?

```{r}
# Solution ----
RunTime <- c(119, 177, 102, 129, 103)
RunTime
```

4.  What code would you use to find the run times of the movies in hours and save them in a vector called `RunTimeHours`? (Hints: Numeric tranformation)

```{r}
# Solution ----
RunTimeHours <- RunTime / 60
RunTimeHours
```

5.  What code would you use to create a data frame named `MovieInfo` containing the vectors created in Problem 1, Problem 2, and Problem 3? (Hints: `data.frame()`)

```{r}
# Solution ----
MovieInfo <- data.frame(Movie, Year, RunTime)
MovieInfo
str(MovieInfo)
```


# Manipulation

## Problems

Suppose we have the following data frame named `colleges` ([download here](data/colleges.xlsx)):

```{r echo=FALSE}
library(rio)
d <- import(file = "data/colleges.xlsx")
knitr::kable(d)
```

1.  What code would you use to select the first, third, tenth, and twelfth entries in the `TopSalary` vector from the `Colleges` data frame? (Hints: Indexing with `[]` operator)

```{r}
# Solution ----
library(rio)
colleges <- import(file = "data/colleges.xlsx")
str(colleges)
colleges$College <- factor(colleges$College) # convert to factor
colleges$TopSalary[c(1, 3, 10, 12)]
```

2.  What code would you use to select the elements of the `MedianSalary` vector where the `TopSalary` is greater than \$400,000? (Hints: `d$MedianSalary[d$TopSalary>400000]`)

```{r}
# Solution ----
colleges$MedianSalary[colleges$TopSalary>400000]
```

3.  What code would you use to select the rows of the data frame for colleges with less than or equal to 1000 employees? (Hints: `d[condition, ]`)

```{r}
# Solution ----
colleges[colleges$Employees<=1000, ]
```


4.  What code would you use to select a sample of 5 colleges from this data frame (there are 14 rows)? (Hints: `d[sample(x = 1:14, size = 5, replace = F),]`)

```{r}
# Solution ----
colleges[sample(x = 1:14, size = 5, replace = F),]
```

Suppose we have the following data frame named Countries ([download here](data/countries.xlsx)):

```{r echo=FALSE}
library(rio)
d <- import(file = "data/countries.xlsx")
knitr::kable(d)
```


5.  What could would you use to select the rows of the data frame that have GDP per capita less than 10000 and are not in the Asia region?

```{r}
# Solution ----
library(rio)
Countries <- import(file = "data/countries.xlsx")
Countries$Region <- factor(Countries$Region)
Countries[Countries$GDPcapita < 10000 & !(Countries$Region %in% "Asia"), ]
```


6.  What code would you use to select a sample of three nations from this data frame (There are 10 rows)?

```{r}
# Solution ----
Countries[sample(x = 1:10, size = 3, replace = F),]
```

7.  What code would you use to select which nations saw a population percent increase greater that 1.5%?


```{r}
# Solution ----
Countries[Countries$PctIncrease> 1.5,]
```


Suppose we have the following data frame named Olympics ([download here](data/olympics.xlsx)):

```{r echo=FALSE}
library(rio)
d <- import(file = "data/olympics.xlsx")
knitr::kable(d)
```

8.  What code would you use to select the rows of the data frame where the host nation was also the medal leader?

```{r}
# Solution ----
library(rio)
Olympics <- import(file = "data/olympics.xlsx")
Olympics$Type <- factor(Olympics$Type)
Olympics$Host <- factor(Olympics$Host)
Olympics$Leader <- factor(Olympics$Leader)
Olympics[as.character(Olympics$Host) == as.character(Olympics$Leader), ]
```


9.  What code would you use to select the rows of the data frame where the number of competitors per event is greater than 35?

```{r}
# Solution ----
Olympics[Olympics$Competitors/Olympics$Events > 35, ]
```


10.  What code would you use to select the rows of the data frame where the number of competing nations in the Winter Olympics is at least 80?

```{r}
# Solution ----
Olympics[Olympics$Nations >= 80 & Olympics$Type == "Winter", ]
```


# Packages

## Problems

1. Install the **Ecdat** package. (Hints: `install.packages()`)

```{r eval=FALSE}
# Solution ----
install.packages("Ecdat")
```

2. Say that we previously installed the **Ecdat** library into R and wanted to call the library to access datasets from it. What code would we use to call the library? (Hints: `library()`)

```{r}
# Solution ----
library("Ecdat")
```


3.  Say that we then wanted to call the dataset `Diamond` from the **Ecdat** library. What code would we use to load this dataset into R? (Hints: `data()`)

```{r}
# Solution ----
data("Diamond")
str(Diamond)
```


# Frequency and numerical exploratory analyses

## Problems

Load the `leuk` dataset from the *MASS* library. This dataset is the survival times (`time`), white blood cell count (`wbc`), and the presence of a morphologic characteristic of white blood cells (`ag`).

1.  Generate the frequency table for the presence of the morphologic characteristic.

```{r}
# Solution ----
data("leuk", package = "MASS")
str(leuk)
table(leuk$ag)
DescTools::Desc(leuk$ag, plotit = F, )
```

2.  Find the median and mean for survival time.

```{r}
# Solution ----
median(leuk$time)
```

3.  Find the range, IQR, variance, and standard deviation for white blood cell count.

```{r}
# Solution ----
diff(range(leuk$wbc)) # range
IQR(leuk$wbc)
var(leuk$wbc)
sd(leuk$wbc)
```


4.  Find the correlation between white blood cell count and survival time.

```{r}
# Solution ----
cor(leuk$wbc, leuk$time)
```


Load the `survey` dataset from the *MASS* library. This dataset contains the survey responses of a class of college students.

5.  Create the contingency table of whether or not the student smoked (`Smoke`) and the student's exercise regimen (`Exer`). (Hints: `table()`, `DescTools::Desc()`)

```{r}
# Solution ----
data("survey", package = "MASS")
str(survey)
# recode factor Smoke
levels(survey$Smoke)
survey$Smoke <- car::recode(survey$Smoke, 'c("Heavy","Occas","Regul")="Yes";"Never"="No"')
table(survey$Smoke, survey$Exer)
DescTools::Desc(Smoke~Exer, data=survey, plotit = F, )
```


6.  Find the mean and median of the student's heart rate (`Pulse`). (Hints: `summary()`, `DescTools::Desc()`, `psych::describe()`)

```{r}
# Solution ----
mean(survey$Pulse, na.rm = T)
median(survey$Pulse, na.rm = T)
summary(survey$Pulse)
psych::describe(survey$Pulse)
DescTools::Desc(survey$Pulse, plotit = F)
```

7.  Find the range, IQR, variance, and standard deviation for student age (`Age`).

```{r}
# Solution ----
diff(range(survey$Age)) # range
IQR(survey$Age)
var(survey$Age)
sd(survey$Age)
```

8.  Find the correlation between the span of the student's writing hand (`Wr.Hnd`) and nonwriting hand (`NW.Hnd`). (Hints: `cor()`,  `DescTools::Desc()`)

```{r}
# Solution ----
cor(survey$Wr.Hnd, survey$NW.Hnd, use = "complete.obs")
DescTools::Desc(Wr.Hnd~NW.Hnd, data=survey, plotit = F)
```

Load the `Housing` dataset from the *Ecdat* library. This dataset looks at the variables that affect the sales price of houses.

9.  Create the contingency table of whether or not the house has a recreation room (`recroom`) and whether or not the house had a full basement (`fullbase`).

```{r}
# Solution ----
data("Housing", package = "Ecdat")
str(Housing)
table(Housing$recroom, Housing$fullbase)
DescTools::Desc(recroom~fullbase, data=Housing, plotit = F, )
```

10.  Find the mean and median of the house's lot size (`lotsize`).

```{r}
# Solution ----
mean(Housing$lotsize)
median(Housing$lotsize)
```

11.  Find the range, IQR, variance, and standard deviation for the sales price (`price`).

```{r}
# Solution ----
DescTools::Desc(Housing$price, plotit = F)
```

12.  Find the correlation between the sales price of the house (`price`) and the number of bedrooms (`bedrooms`).

```{r}
# Solution ----
cor(Housing$price, Housing$bedrooms)
DescTools::Desc(price~bedrooms, data=Housing, plotit = F)
```



# Graphical exploratory analyses

Load the `Star` dataset from the *Ecdat* library. This dataset looks at the affect on class sizes on student learning.

1.  Generate the scatterplot of the student's math score `tmathssk` and reading score `treadssk`. (Hints: `plot()`, `ggplot() + geom_point()`)

```{r}
# Solution ----
data("Star", package = "Ecdat")
str(Star)
```


```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
plot(Star$tmathssk, Star$treadssk)
```

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
library(ggplot2)
ggplot(data = Star, mapping = aes(x=tmathssk, y=treadssk)) + geom_point()
```

2.  Generate the histogram of the years of teaching experience `totexpk`. (Hints: `hist()`, `ggplot() + geom_histogram()`)

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
hist(Star$totexpk)
```

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
library(ggplot2)
ggplot(data = Star, mapping = aes(x=totexpk)) + geom_histogram(binwidth = 2, fill="grey", col="blue")
```

3.  Create a new variable in the `Star` dataset called `totalscore` that is the sum of the student's math score `tmathssk` and reading score `treadssk`. (Hints: tranformation)

```{r}
# Solution ----
Star$totalscore <- Star$tmathssk + Star$treadssk
```


4.  Generate a boxplot of the student's total score `totalscore` split out by the class size type `classk`. (Hints: `boxplot()`, `ggplot() + geom_boxplot()`)

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
boxplot(totalscore~classk, data = Star)
```

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
library(ggplot2)
ggplot(data = Star, mapping = aes(x=classk, y=totalscore)) + geom_boxplot()
```


Load the `survey` dataset from the *MASS* library. This dataset contains the survey responses of a class of college students.

5.  Generate the scatterplot of the student's height `Height` and writing hand span `Wr.Hnd`.

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
plot(survey$Height, survey$Wr.Hnd)
```

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
library(ggplot2)
ggplot(data = survey, mapping = aes(x=Height, y=Wr.Hnd)) + geom_point()
```


6.  Generate the histogram of student age `Age`.

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
hist(survey$Age)
```

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
library(ggplot2)
ggplot(data = survey, mapping = aes(x=Age)) + geom_histogram(binwidth = 10, fill="grey", col="blue", boundary=10) + labs(x="Age", y="Count")
```


7.  Generate a boxplot of the student's heart rate `Pulse` split out by the student's exercise regimen `Exer`.

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
boxplot(Pulse~Exer, data = survey)
```

```{r fig.margin=F, fig.width=4, fig.height=4}
# Solution ----
library(ggplot2)
ggplot(data = survey, mapping = aes(x=Exer, y=Pulse, fill=Exer)) + geom_boxplot() + theme_bw() + theme(legend.position = "none")
```

# Reference

